---
title: "Nambari Day Three (Heart Disease)"
author: "Lumumba Wandera Victor"
date: "2023-07-05"
output:
  html_document: default
  word_document: default
  pdf_document: default
---
### Set up Rstudio
Setting up RMarkdown when opening it enables you to create dynamic, reproducible, and visually appealing reports, presentations, and documents, that can help you communicate your data analysis and research findings more effectively.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE,comment = NA, message=FALSE,
                      fig.height=4, fig.width=6)
```


### Set up the work directory
```{r}
setwd("/Users/user/Desktop/Biostatistics WorkShop/")
```


### Load the following libraries
```{r}
library(ISLR2)
library(MASS)
library(caret)
library(splines)
library(pROC)
library(rattle)
```

### HEART DISEASE DATA
```{r}
HeartData <- read.csv("Heart.csv", header=TRUE)
HeartData <- na.omit(HeartData)
```

### Attach the dataset
```{r}
attach(HeartData)
```

### View the dataset
```{r}
head(HeartData,10)
```

### View the Boxplots
```{r}
par(mfrow=c(1,4))
boxplot(Age ~ as.factor(AHD))
boxplot(MaxHR ~ as.factor(AHD))
boxplot(Chol ~ as.factor(AHD))
boxplot(Age ~ as.factor(AHD))
```

```{r}
par(mfrow=c(1,1))
pairs( cbind( Chol, MaxHR, RestBP,Age), pch=19, lower.panel=NULL, cex=.5)
HeartData$HD = as.factor(AHD)
```

### looking at classification based on p.hat = .5 cutoff
### 10-fold CV, repeated 5 times
```{r}
train_model <- trainControl(method = "repeatedcv", number = 5, repeats=10)


model.cart <- train(
  HD ~ Age + as.factor(Sex) + as.factor(Thal)
        + Chol + MaxHR + RestBP + Fbs + RestECG + ExAng, 
  data = HeartData, 
  method = "rpart",
  trControl = train_model)

model.cart

model.cart$finalModel

confusionMatrix(predict(model.cart, HeartData), 
                reference=HeartData$HD, positive="Yes")
```

### summary(model.cart$finalModel)
```{r}
fancyRpartPlot(model.cart$finalModel)
```

### cart model for under 50 vs over 50
### train on everyone
```{r}
model.cartl50 <- train(
  HD ~ Age + as.factor(Sex) + as.factor(Thal)
  + Chol + MaxHR + RestBP + Fbs + RestECG + ExAng, 
  data = HeartData, 
  method = "rpart",
  trControl = train_model)
```

### Below we will examine the performance on the two age groups. Is there a disparity? In what cases does this matter?

### predict on under 50
```{r}
  confusionMatrix(predict(model.cartl50, HeartData[HeartData$Age<50,]), 
                reference=HeartData[HeartData$Age<50,]$HD, positive="Yes")
```

### predict on over 50
```{r}
confusionMatrix(predict(model.cartl50, HeartData[HeartData$Age>=50,]), 
                reference=HeartData[HeartData$Age>=50,]$HD, positive="Yes")
```

### To unpack why there is a difference, let's look at variable distribution differences by age
```{r}
par(mfrow=c(1,4))
boxplot(MaxHR[HeartData$Age<50] ~ as.factor(AHD[HeartData$Age<50]),ylim=c(80,200))
boxplot(MaxHR[HeartData$Age>=50] ~ as.factor(AHD[HeartData$Age>=50]),ylim=c(80,200))
boxplot(Chol[HeartData$Age<50] ~ as.factor(AHD[HeartData$Age<50]),ylim=c(100,400))
boxplot(Chol[HeartData$Age>=50] ~ as.factor(AHD[HeartData$Age>=50]),ylim=c(100,400))
```

### categorical variable distribution differences
```{r}
par(mfrow=c(1,2))
plot(as.factor(Thal[HeartData$Age<50]), xlab="Less than 50", ylab="Count")
plot(as.factor(Thal[HeartData$Age>=50]), xlab="50 and older", ylab="Count")
```

```{r}
detach(HeartData)
```



