---
title: "Malaria Infection Cases Simulation"
author: "Lumumba Wandera Victor"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Simulate Malaria Data
```{r}
# Load necessary libraries
library(dplyr)
library(tibble)
library(MASS)

# Set the seed for reproducibility
set.seed(123)

# Number of observations
n <- 500

# Simulate continuous independent variables
continuous_vars <- tibble(
  age = rnorm(n, mean = 35, sd = 12),
  income = rnorm(n, mean = 30000, sd = 15000),
  temperature = rnorm(n, mean = 25, sd = 5),
  rainfall = rnorm(n, mean = 100, sd = 50),
  mosquito_density = rnorm(n, mean = 20, sd = 10),
  distance_to_health_facility = rnorm(n, mean = 10, sd = 5),
  bed_net_usage = rnorm(n, mean = 50, sd = 20),
  education_years = rnorm(n, mean = 10, sd = 4),
  household_size = rnorm(n, mean = 5, sd = 2),
  urbanization_rate = rnorm(n, mean = 70, sd = 15)
)

# Simulate categorical independent variables
categorical_vars <- tibble(
  gender = sample(c("male", "female"), n, replace = TRUE),
  occupation = sample(c("farmer", "teacher", "merchant", "unemployed"), n, replace = TRUE),
  housing_quality = sample(c("poor", "average", "good"), n, replace = TRUE),
  access_to_clean_water = sample(c("yes", "no"), n, replace = TRUE),
  sanitation_facilities = sample(c("poor", "average", "good"), n, replace = TRUE),
  region = sample(c("north", "south", "east", "west"), n, replace = TRUE),
  vaccination_status = sample(c("yes", "no"), n, replace = TRUE),
  insecticide_treated_nets = sample(c("yes", "no"), n, replace = TRUE),
  malaria_history = sample(c("yes", "no"), n, replace = TRUE),
  health_insurance = sample(c("yes", "no"), n, replace = TRUE)
)

# Combine continuous and categorical variables
independent_vars <- bind_cols(continuous_vars, categorical_vars)

# Generate the outcome variable
# Create a model matrix for the categorical variables
cat_model_matrix <- model.matrix(~ gender + occupation + housing_quality + access_to_clean_water +
                                   sanitation_facilities + region + vaccination_status +
                                   insecticide_treated_nets + malaria_history + health_insurance, 
                                 data = categorical_vars)[, -1]

# Combine continuous variables and the model matrix for categorical variables
combined_vars <- cbind(as.matrix(continuous_vars), cat_model_matrix)

# Create random coefficients for the variables
beta <- rnorm(ncol(combined_vars))

intercept <- 10000  # Adjust this value to increase/decrease infection rate

# Calculate the linear predictor
linear_predictor <- combined_vars %*% beta + intercept

# Generate probabilities using the logistic function
prob <- 1 / (1 + exp(-linear_predictor))

# Generate the binary outcome variable
outcome <- rbinom(n, 1, prob)

# Combine the outcome variable with the independent variables
malaria_data <- bind_cols(
  tibble(infected = outcome),
  independent_vars
)

# Display the first few rows of the dataset
print(head(malaria_data))
```

This sequence of steps simulates the outcome variable (infected) for a dataset by first creating a set of predictors, assigning random coefficients to them, and then combining these predictors into a linear predictor. The logistic function converts this linear predictor into probabilities, which are then used to generate binary outcomes through a Bernoulli process. By setting a high intercept value, we ensure that the base probability of infection is high, resulting in more infected cases in the dataset. Adjusting the intercept value allows us to control the prevalence of the outcome.











