---
title: "SIMULATION OF MALARIA DATA"
author: "Lumumba Wandera Victor"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, message = FALSE, warning = FALSE,
                      fig.height = 4, fig.width = 5)
```

### Load the Data
```{r}
library(caret) #for machine learning models
library(psych) ##for description of  data
library(ggplot2) ##for data visualization
library(caretEnsemble)##enables the creation of ensemble models
library(tidyverse) ##for data manipulation
library(mlbench)  ## for benchmarking ML Models
library(flextable) ## to create and style tables
library(mltools) #for hyperparameter tuning
library(tictoc) #for determining the time taken for a model to run
library(ROSE)  ## for random oversampling
library(smotefamily) ## for smote sampling
library(ROCR) ##For ROC curve
```

### Import the data
```{r}
malaria_data <- read.csv("malaria.simulateddata.csv")
head(malaria_data,10)
```
Explanation of the Variables:

### Continuous Variables:

age: Age of the individual.
income: Annual income of the individual.
temperature: Average temperature of the area.
rainfall: Average rainfall in the area.
mosquito_density: Density of mosquitoes in the area.
distance_to_health_facility: Distance to the nearest health facility.
bed_net_usage: Percentage of bed net usage in the household.
education_years: Number of years of education.
household_size: Number of people in the household.
urbanization_rate: Percentage of urbanization in the area.

### Categorical Variables:

gender: Gender of the individual (male or female).
occupation: Occupation of the individual.
housing_quality: Quality of housing (poor, average, or good).
access_to_clean_water: Access to clean water (yes or no).
sanitation_facilities: Quality of sanitation facilities (poor, average, or good).
region: Geographical region (north, south, east, or west).
vaccination_status: Whether the individual is vaccinated (yes or no).
insecticide_treated_nets: Whether insecticide-treated nets are used (yes or no).
malaria_history: History of malaria infection (yes or no).
health_insurance: Whether the individual has health insurance (yes or no).


### Convert All Character Variables to Factors
```{r}
factor_cols <- sapply(malaria_data, is.character)
malaria_data[, factor_cols] <- lapply(malaria_data[, factor_cols], as.factor)
head(malaria_data,5)
```

### Create Factors for the Outcome variable
```{r}
malaria_data$infected <- factor(malaria_data$infected, levels = c(0,1),
                                labels = c("Not Infected", "Infected"))
head(malaria_data,5)
```

### Dimension of the Data
```{r}
dim(malaria_data)
```

### View the Variable Names
```{r}
names(malaria_data)
```
### Explore the data
```{r}
attach(malaria_data)
summary(malaria_data) ###Descriptive Statistics
describe(malaria_data)###Descriptive Statistics
sum(is.na(malaria_data))###Check for missing data
```

### Bag Graph of the Predicted variable
```{r, fig.height=5, fig.width=7}
Obj <- malaria_data %>%
  group_by(infected)%>%
  summarize(Percentage = n()/nrow(malaria_data)*100)

ggplot(Obj, aes(x= infected, y = Percentage, fill= infected))+
  geom_bar(stat = "identity")+
  geom_text(aes(label = sprintf("%.1f%%",Percentage)),vjust=-0.2)+
  labs(title = "",
       x ="Malaria Detected",
       y="Percent")+
   theme(axis.text.x = element_text(angle = 0, vjust = 0.5))+
  ggtitle("Malaria Incidence Detected")
```


### DATA PARTITION FOR MACHINE LEARNING
```{r}
head(malaria_data)
str(malaria_data)
```

```{r}
set.seed(123)
ind=sample(2, nrow(malaria_data),replace =T, prob=c(0.70,0.30))
train=malaria_data[ind==1,]
test= malaria_data[ind==2,]
```

## Get the dimensions of your train and test data
```{r}
dim(train)
dim(test)
```

Now Let's train some machine learning models using package caret. The caret R package (Kuhn et al. 2021) (short for Classification And REgression Training) to carry out machine learning tasks in RStudio. The caret package offers a range of tools and models for classification and regression machine learning problems. In fact, it offers over 200 different machine learning models from which to choose.  Don’t worry, we don’t expect you to use them all!

### VIEW THE MODELS IN CARET
```{r}
models= getModelInfo()
names(models)
```

### Check for zero variance predictors:
```{r}
nzv <- nearZeroVar(malaria_data[,-1], saveMetrics = TRUE)
```

```{r}
print(nzv)
```

The results above show that there is no feature with zero variance
## Remove nzv
```{r}
#mdata1 <- mdata[, !nzv$nzv]
#dim(mdata1)
```

```{r}
missing_values <- colSums(is.na(train))
missing_values
```



### Prepare training scheme for cross-validation
```{r}
control <- trainControl(method="repeatedcv", number=10, repeats=5)
```

###TRAIN YOUR ML MODELS
# Train a SVM model
```{r}
set.seed(123)
tic()
SvmModel <- train(infected~.,data=train, method="svmRadial", trControl=control, na.action = na.omit)
toc()
SvmModel
```

### Plot the Model (Accuracy and Cost)
```{r}
results <- SvmModel$results
plot(SvmModel)
```

### Plot the Accuracy and Sigma
```{r}
results
```

### Accuracy and Sigma
```{r}
ggplot(results, aes(x = sigma, y = Accuracy)) +
  geom_line() +
  geom_point() +
  labs(title = "Accuracy vs. Sigma for SVM Model",
       x = "Sigma",
       y = "Accuracy")
```


### Accuracy and Kappa Value
```{r}
ggplot(results, aes(x = Kappa, y = Accuracy)) +
  geom_line() +
  geom_point() +
  labs(title = "Accuracy vs. Kappa for SVM Model",
       x = "Kappa",
       y = "Accuracy")

```

### Prediction using the Test Set
```{r}
Svmpred= predict(SvmModel,newdata = test)
SVM.cM<- confusionMatrix(Svmpred,as.factor(test$infected), positive = 'Infected', mode='everything')
SVM.cM
```

```{r}
M1<- SVM.cM$byClass[c(1, 2, 5, 7, 11)]
M1
```



### View the Predicted Case and the Ground Truth
```{r}
data.f <- data.frame(Svmpred, test$infected)
head(data.f,5)
```

### plotting confusion matrix
```{r}
SVM.cM$table
fourfoldplot(SVM.cM$table, col=c(2,3), main="Imbalanced SVM Confusion Matrix")
plot(varImp(SvmModel, scale=T))
```

## Train the decision Tree
```{r}
set.seed(123)
tic()
DTModel <- train(infected~., data=train, method="rpart", trControl=control)
toc()
DTModel
```

```{r}
DTpred=predict(DTModel,newdata = test)
DT.cM<- confusionMatrix(DTpred,
                        reference = test$infected, positive = 'Infected', mode='everything')
DT.cM
```

### Prediction
```{r}
M2<- DT.cM$byClass[c(1, 2, 5, 7, 11)]
M2
```

### plotting confusion matrix
```{R}
DT.cM$table
varImp(DTModel)
plot(varImp(DTModel))
fourfoldplot(DT.cM$table, col=rainbow(4), main="Imbalanced RF Confusion Matrix")
```


## Train a Random Forest model
```{r}
set.seed(123)
tic()
RFModel <- train(infected~., data=train, method="rf", trControl=control)
toc()
RFModel
```

### Prediction
```{r}
RFpred=predict(RFModel,newdata = test)
RF.cM<- confusionMatrix(RFpred,
                        reference = test$infected , positive = 'Infected', mode='everything')
RF.cM
```

```{R}
M3<- RF.cM$byClass[c(1, 2, 5, 7, 11)]
M3
```

### plotting confusion matrix
```{R}
RF.cM$table
fourfoldplot(RF.cM$table, col=rainbow(4), main="Imbalanced RF Confusion Matrix")
varImp(RFModel)
plot(varImp(RFModel, scale=T))
```

### Calculate the and Plot the ROC
```{r, fig.height=4, fig.width=6}
library(pROC)
library(ggplot2)

# Compute the ROC curve
ROC <- roc(response = test$infected,
           predictor = predict(RFModel, newdata = test, type = 'prob')$Infected)

# Extract the AUC
auc_value <- auc(ROC)

# Convert ROC curve to a data frame
roc_data <- data.frame(
  tpr = ROC$sensitivities,
  fpr = 1 - ROC$specificities
)

# Plot the ROC curve using ggplot2
ggplot(roc_data, aes(x = fpr, y = tpr)) +
  geom_line(color = "blue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "The ROC Curve and the Area Under the Curve",
    x = "False Positive Rate",
    y = "True Positive Rate"
  ) +
  annotate("text", x = 0.5, y = 0.5, label = paste("AUC =", round(auc_value, 4)), size = 5, hjust = 0)

```


### Train an Logisitic Regression model
```{r}
set.seed(123)
lrModel <- train(infected~., 
                 data=train, 
                 method="glm", 
                 trControl=control)
lrModel
```

### Prediction
```{r}
lrpred=predict(lrModel,newdata = test)
lr.cM<- confusionMatrix(lrpred,
                        reference = test$infected, positive = 'Infected', mode='everything')
lr.cM
```

```{r}
M4<- lr.cM$byClass[c(1, 2, 5, 7, 11)]
M4
```

### plotting confusion matrix
```{r}
lr.cM$table
fourfoldplot(lr.cM$table, col=rainbow(4), main="Imbalanced LR Confusion Matrix")
plot(varImp(lrModel, scale=T))
```

```{r}
# Compute the ROC curve
ROC <- roc(response = test$infected,
           predictor = predict(lrModel, newdata = test, type = 'prob')$Infected)

# Extract the AUC
auc_value <- auc(ROC)

# Convert ROC curve to a data frame
roc_data <- data.frame(
  tpr = ROC$sensitivities,
  fpr = 1 - ROC$specificities
)

# Plot the ROC curve using ggplot2
ggplot(roc_data, aes(x = fpr, y = tpr)) +
  geom_line(color = "blue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "The ROC Curve and the Area Under the Curve",
    x = "False Positive Rate",
    y = "True Positive Rate"
  ) +
  annotate("text", x = 0.5, y = 0.5, label = paste("AUC =", round(auc_value, 4)), size = 5, hjust = 0)
```

### Train a k- Nearest Neigbour model
```{r}
set.seed(123)
knnModel <- train(infected~., 
                  data=train, 
                  method="knn", 
                  trControl=control)
knnModel
```

### View the Nearest Neighbors
```{r}
knnModel$finalModel
```

### Prediction
```{r}
knnpred=predict(knnModel,newdata = test)
knn.cM<- confusionMatrix(knnpred,as.factor(test$severe_maleria), positive = 'Infected', mode='everything')
knn.cM
m4<- knn.cM$byClass[c(1, 2, 5, 7, 11)]
m4
```

### plotting confusion matrix
```{r}
knn.cM$table
fourfoldplot(knn.cM$table, col=rainbow(4), main="Imbalanced KNN Confusion Matrix")
plot(varImp(knnModel, scale=T))
```



### Alternative KNN Approach
```{r}
x <- train[,-18]
  y <- train[,18]
```

```{r}
library(e1071)
obj2 <- tune.knn(x, y, k = 1:10, tunecontrol = tune.control(sampling = "boot"))
summary(obj2)
plot(obj2)
```

```{r}
fit_knn <- knn3(factor(severe_maleria) ~ ., data = train, k = 51)
fit_knn$k
```


## Quadratic and Linear Discriminant Analysis
```{r, warning=FALSE}
qda_fit <- qda(factor(severe_maleria) ~., data = train)
qda_preds <- predict(qda_fit, test)
```


```{r, warning=FALSE}
confusionMatrix(data = as.factor(qda_preds$class), reference = as.factor(test$severe_maleria))
```

## Lda
```{r}
lda_fit <- lda(factor(severe_maleria) ~., data = train)
lda_preds <- predict(lda_fit, test)
confusionMatrix(data = as.factor(lda_preds$class), reference = as.factor(test$severe_maleria))
```


## ROC

```{r, message=FALSE, warning=FALSE}
pred_qda <- qda_preds$posterior[,2]
test %>% mutate(pred=pred_qda, label=as.factor(severe_maleria)) %>%
  ggplot(aes(label,pred)) + geom_boxplot()
```


```{r, message=FALSE, warning=FALSE}
library(pROC)

roc_qda <- roc(test$severe_maleria, pred_qda)
plot(roc_qda)
```


We can also create a nicer looking plot using `ggroc`:

```{r}
ggroc(list("QDA" = roc_qda)) +
  theme(legend.title = element_blank()) +
  geom_segment(aes(x = 1, xend = 0, y = 0, yend = 1), color = "black", linetype = "dashed") +
  xlab("Specificity") +
  ylab("Sensitivity") 
```


Here are the results for LDA

```{r}
pred_lda <- lda_preds$posterior[,2]
roc_lda <- roc(test$severe_maleria, pred_lda)

ggroc(list("QDA" = roc_qda, "LDA" = roc_lda)) +
  theme(legend.title = element_blank()) +
  geom_segment(aes(x = 1, xend = 0, y = 0, yend = 1), color = "black", linetype = "dashed") +
  xlab("Specificity") +
  ylab("Sensitivity") 
```

We can also compare to KNN with two different values of `k`: 5 and 51.

```{r}
fit1 <- knn3(factor(severe_maleria)~., data = train, k = 5)

pred_knn <- predict(fit1, newdata = test)[,2]
roc_knn <- roc(test$severe_maleria, pred_knn)


fit2 <- knn3(factor(severe_maleria)~., data = train, k = 51)
pred_knn2 <- predict(fit2, newdata = test)[,2]
roc_knn2  <- roc(test$severe_maleria, pred_knn2)

ggroc(list("QDA" = roc_qda, "LDA" = roc_lda, "kNN, k = 5" = roc_knn2, "kNN, k = 51" = roc_knn)) +
  theme(legend.title = element_blank()) +
  geom_segment(aes(x = 1, xend = 0, y = 0, yend = 1), color = "black", linetype = "dashed") +
  xlab("Specificity") +
  ylab("Sensitivity")
```

To summarize these curves into one single number that can be compared across methods, it is common to take the area under the ROC curve (abbreviated AUC or AUROC). Higher values indicate better performance.

```{r}
auc(roc_qda)
auc(roc_lda)
auc(roc_knn)
auc(roc_knn2)
```

### Train a Neural Net model
```{r}
set.seed(123)
nnModel <- train(factor(severe_maleria)~., 
                 data=train, 
                 method="nnet", 
                 trControl=control)
nnModel
```

### Neural Network Model
```{r}
nnModel$finalModel
```

```{r}
nnpred=predict(nnModel,newdata = test)
nn.cM<- confusionMatrix(nnpred,as.factor(test$severe_maleria), positive = 'Infected', mode='everything')
nn.cM
m5<- nn.cM$byClass[c(1, 2, 5, 7, 11)]
m5
```
### plotting confusion matrix
```{r}
nn.cM$table
fourfoldplot(nn.cM$table, col=rainbow(4), main="Imbalanced NN Confusion Matrix")
plot(varImp(nnModel, scale=T))
```

# Train a Naive Bayes model
```{r}
set.seed(123)
nbModel <- train(factor(severe_maleria)~., data=train, method="nb", trControl=control)
nbModel
```

### Prediction
```{r}
nbpred=predict(nbModel,newdata = test)
nb.cM<- confusionMatrix(nbpred,as.factor(test$severe_maleria), positive = 'Infected', mode='everything')
nb.cM
m6<- nb.cM$byClass[c(1, 2, 5, 7, 11)]
m6
```

### plotting confusion matrix
```{r}
nb.cM$table
fourfoldplot(nb.cM$table, col=rainbow(4), main="Imbalanced NB Confusion Matrix")
plot(varImp(nbModel, scale=T))
```

### Train a Linear Discriminant Analysis model
```{r}
set.seed(123)
ldaModel <- train(factor(severe_maleria)~., data=train, method="lda", trControl=control)
ldaModel
ldapred=predict(ldaModel,newdata = test)
lda.cM<- confusionMatrix(ldapred,as.factor(test$severe_maleria), positive = 'Infected', mode='everything')
m7<- lda.cM$byClass[c(1, 2, 5, 7, 11)]
m7
```

### plotting confusion matrix
```{r}
lda.cM$table
fourfoldplot(lda.cM$table, col=rainbow(4), main="Imbalanced LDA Confusion Matrix")
plot(varImp(ldaModel, scale=T))
```

### Train a Linear Vector Quantization model
```{r}
set.seed(123)
lvqModel <- train(factor(severe_maleria)~., data=train, method="lvq", trControl=control)
lvqModel
lvqpred=predict(lvqModel,newdata = test)
lvq.cM<- confusionMatrix(lvqpred,as.factor(test$severe_maleria), positive = 'Infected', mode='everything')
m7<- lvq.cM$byClass[c(1, 2, 5, 7, 11)]
m7
```

### plotting confusion matrix
```{r}
lvq.cM$table
fourfoldplot(lvq.cM$table, col=rainbow(4), main="Imbalanced LDA Confusion Matrix")
plot(varImp(lvqModel, scale=T))
```

### Train a Bagging model
```{r}
set.seed(123)
bagModel <- train(factor(severe_maleria)~., data=train, method="treebag", trControl=control)
bagModel
bagpred=predict(bagModel,newdata = test)
bag.cM<- confusionMatrix(bagpred,as.factor(test$severe_maleria), positive = 'Infected', mode='everything')
m8<- bag.cM$byClass[c(1, 2, 5, 7, 11)]
m8
```

### plotting confusion matrix
```{r}
bag.cM$table
fourfoldplot(bag.cM$table, col=rainbow(4), main="Imbalanced Bagging Confusion Matrix")
plot(varImp(bagModel, scale=T))
```

### Train a Boosting model
```{r}
set.seed(123)
boModel <- train(factor(severe_maleria)~., data=train, method="ada", trControl=control)
boModel
bopred=predict(boModel,newdata = test)
bo.cM<- confusionMatrix(bopred,as.factor(test$severe_maleria), positive = 'Infected', mode='everything')
m9<- bo.cM$byClass[c(1, 2, 5, 7, 11)]
m9
```

```{r}
bo.cM
```

### plotting confusion matrix
```{r}
bo.cM$table
fourfoldplot(bo.cM$table, col=rainbow(4), main="Imbalanced Boosting Confusion Matrix")
plot(varImp(boModel, scale=T))
```

## TABULATE YOUR RESULTS

```{r}
measure <-round(data.frame(SVM= m1, RF= m2, LR = m3, KNN=m4, NN=m5, NB=m6, Bagging = m8, Boosting = m9), 3)
dim(measure)
rownames(measure)=c('Sensitivity', 'Specificity', 'Precision','F1-Score', 'Balanced Accuracy')
flextable(measure)
measure
View(measure)
#xtable(measure.score, digits = 3)
```

### collect all resamples and compare
```{r}
results <- resamples(list(SVM=SvmModel, Bagging=bagModel,LR=lrModel,NB=nbModel, RF=RFModel))
```

### summarize the distributions of the results 
```{r}
summary(results)
```

### boxplots of results
```{r}
bwplot(results)
```

### dot plots of results
```{r}
dotplot(results)
```

### Oversampling 
# Oversampled data --------------------------------------------------------
```{r}
 over <- ovun.sample(factor(severe_maleria)~., data = train, method = "over")$data
 over
```

```{r}
library(xtable)
xtable(table(over$class))
```
 
## Model building ----------------------------------------------------------
### prepare training scheme for cross-validation
```{r} 
control <- trainControl(method="repeatedcv", number=10, repeats=5)
```
 
## train an SVM model
```{r}
 set.seed(123)
 over.svmModel <- train(factor(severe_maleria)~., data=over, method="svmRadial", trControl=control)
 over.svmModel
 over.svmpred=predict(over.svmModel,newdata = test)
 over.SVM.cM<- confusionMatrix(over.svmpred,as.factor(test$severe_maleria), positive = 'Infected', mode='everything')
 over.SVM.cM
 over.m1<- over.SVM.cM$byClass[c(1, 2, 5, 7, 11)]
 over.m1
```

###plotting confusion matrix
```{r, fig.height=5, fig.width=7}
 over.SVM.cM$table
 fourfoldplot(over.SVM.cM$table, col=rainbow(4), main="Oversampled SVM Confusion Matrix")
```













