---
title: "Network Analysis with R and iGraph"
author: "Lumumba Wandera Victor"
date: "2023-07-11"
output: html_document
---

```{r echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE,comment = NA, message=FALSE,
                      fig.height=4, fig.width=6)
```

## CONTENTS
*1. A quick R introduction/refresher*
*2. Networks in igraph*
*3. Reading network data from files*
*4. Turning networks into igraph objects*
*5. Plotting networks with igraph*
*6. Network and node descriptives*
*7. Distances and paths*
*8. Subgroups and communities*
*9. Assortativity and Homophily*

Install the package "igraph" if you don't have its latest version (1.0.1) The package (www.igraph.org) is maintained by Gabor Csardi and Tamas Nepusz.

### Install and load the following library
```{r}
library(igraph)
```

#### 1. A quick R introduction/reminder
You can assign a value to an object using assign(),  "<-", or "=".
```{r}
x <- 3         # Assignment
x              # Evaluate the expression and print result

y <- 4         # Assignment
y + 5          # Evaluation, y remains 4

z <- x + 17*y  # Assignment
z              # Evaluation

rm(z)          # Remove z: deletes the object.
z              # Error!
```

### Value comparisons:
Comparisons return boolean values: TRUE or FALSE (often abbreviated to T and F)

```{r}
2==2  # Equality
2!=2  # Inequality
x <= y # less than or equal: "<", ">", and ">=" also work
```

### Special constants
####  NA, NULL, Inf, -Inf, NaN
##### NA - missing or undefined data

```{r}
5 + NA      # When used in an expression, the result is generally NA
is.na(5+NA) # Check if missing

# NULL - an empty object, e.g. a null/empty list
10 + NULL     # use returns an empty object (length zero)
is.null(NULL) # check if NULL

# Inf and -Inf represent positive and negative infinity
# They can be returned by  mathematical operations like division of a number by zero:

5/0
is.finite(5/0) # Check if a number is finite

# NaN (Not a Number) - the result of an operation that cannot be reasonably defined 
0/0
is.nan(0/0)
```

### Vectors 
````{r}
v1 <- c(1, 5, 11, 33)       # Numeric vector, length 4
v2 <- c("hello","world")    # Character vector, length 2 (a vector of strings)
v3 <- c(TRUE, TRUE, FALSE)  # Logical vector, same as c(T, T, F)
````

### Combining different types of elements in one vector will coerce the elements to the least restrictive type:

````{r}
v4 <- c(v1,v2,v3,"boo") 	# All elements turn into strings
v4
````

### Other ways to create vectors:
````{r}
v <- 1:7         # same as c(1,2,3,4,5,6,7)  
v <- rep(0, 77)  # repeat zero 77 times: v is a vector of 77 zeroes
v <- rep(1:3, times=2) # Repeat 1,2,3 twice  
v <- rep(1:10, each=2) # Repeat each element twice  
v <- seq(10,20,2) # sequence: numbers between 10 and 20, in jumps of 2  
v
````

````{r}
length(v)        # check the length of the vector

v1 <- 1:5         # 1,2,3,4,5
v2 <- rep(1,5)    # 1,1,1,1,1 
````

### Element-wise operations:
````{r}
v1 + v2      # Element-wise addition
v1 + 1       # Add 1 to each element
v1 * 2       # Multiply each element by 2
v1 + c(1,7)  # This doesn't work: (1,7) is a vector of different length
`````

### Mathematical operations:
````{r}
sum(v1)      # The sum of all elements
mean(v1)     # The average of all elements
sd(v1)       # The standard deviation
cor(v1,v1*5) # Correlation between v1 and v1*5 
```

#### Logical operations:
````{r}
v1 > 2       # Each element is compared to 2, returns logical vector
v1==v2       # Are corresponding elements equivalent, returns logical vector.
v1!=v2       # Are corresponding elements *not* equivalent? Same as !(v1==v2)
(v1>2) | (v2>0)   # | is the boolean OR, returns a vector.
(v1>2) & (v2>0)   # & is the boolean AND, returns a vector.
(v1>2) || (v2>0)  # || is the boolean OR, returns a single value
(v1>2) && (v2>0)  # && is the boolean AND, ditto
````

### Vector elements
````{r}
v1[3]             # third element of v1
v1[2:4]           # elements 2, 3, 4 of v1
v1[c(1,3)]        # elements 1 and 3 - note that your indexes are a vector
v1[c(T,T,F,F,F)]  # elements 1 and 2 - only the ones that are TRUE
v1[v1>3]          # v1>3 is a logical vector TRUE for elements >3
````

### NOTE: If you are used to languages indexing from 0, R will surprise you by indexing from 1.

###### To add more elements to a vector, simply assign them values.
`````{r}
v1[6:10] <- 6:10
````

### We can also directly assign the vector a length:
````{r}
length(v1) <- 15 # the last 5 elements are added as missing data: NA
```

### Factors
#### Factors are used to store categorical data.

````{r}
eye.col.v <- c("brown", "green", "brown", "blue", "blue", "blue")         #vector
eye.col.f <- factor(c("brown", "green", "brown", "blue", "blue", "blue")) #factor
eye.col.v
eye.col.f
````

R will identify the different levels of the factor - e.g. all distinct values. The data is stored internally as integers - each number corresponding to a factor level.

```{r}
levels(eye.col.f)  # The levels (distinct values) of the factor (categorical variable)

as.numeric(eye.col.f)  # The factor as numeric values: 1 is  blue, 2 is brown, 3 is green
as.numeric(eye.col.v)  # The character vector, however, can not be coerced to numeric

as.character(eye.col.f)  
as.character(eye.col.v)
```

### Matrces & Arrays
#####  A matrix is a vector with dimensions:
````{r}
m <- rep(1, 20)   # A vector of 20 elements, all 1
dim(m) <- c(5,4)  # Dimensions set to 5 & 4, so m is now a 5x4 matrix
m
````

### Create a matrix using matrix():
````{r}
m <- matrix(data=1, nrow=5, ncol=4)  # same matrix as above, 5x4, full of 1s
m <- matrix(1,5,4) 			             # same matrix as above
dim(m)                               # What are the dimensions of m?
````

### Create a matrix by combining vectors:
```{r}
m <- cbind(1:5, 5:1, 5:9)  # Bind 3 vectors as columns, 5x3 matrix
m <- rbind(1:5, 5:1, 5:9)  # Bind 3 vectors as rows, 3x5 matrix

m <- matrix(1:10,10,10)
m
```

### Select matrix elements: 
````{r}
m[2,3]  # Matrix m, row 2, column 3 - a single cell
m[2,]   # The whole second row of m as a vector
m[,2]   # The whole second column of m as a vector
m[1:2,4:6] # submatrix: rows 1 and 2, columns 4, 5 and 6
m[-1,]     # all rows *except* the first one

m[1,]==m[,1]  # Are elements in row 1 equivalent to corresponding elements from column 1? 
m>3           # A logical matrix: TRUE for m elements >3, FALSE otherwise
m[m>3]        # Selects only TRUE elements - that is ones greater than 3
```

```{r}
t(m)          # Transpose m     
m <- t(m)     # Assign m the transposed m
m %*% t(m)    # %*% does matrix multiplication
m * m         # * does element-wise multiplication
```

### Arrays: more than 2 dimensions
##### Created with the array() function:
`````{r}
a <- array(data=1:18,dim=c(3,3,2)) # 3d with dimensions 3x3x2
a <- array(1:18,c(3,3,2))          # the same array
a
```

### Lists 
Lists are collections of objects (e.g. of strings, vectors, matrices, other lists, etc.)

````{r}
l1 <- list(boo=v1,foo=v2,moo=v3,zoo="Animals!")  # A list with four components
l2 <- list(v1,v2,v3,"Animals!")

l3 <- list()
l4 <- NULL

l1["boo"]      # Access boo: this returns a list.
l1[["boo"]]    # Access boo: this returns the numeric vector
l1[[1]]        # Returns the first component of the list, equivalent to above.
l1$boo         # Named elements can be accessed using the $ operator - equivalent to [[]]
```
### Add more elements to a list:
````{r}
l3[[1]] <- 11 # add an element to the empty list l3
l4[[3]] <- c(22, 23) # add a vector as element 3 in the empty list l4. 
                     # Since we added element 3, elements 1 & 2 will be generated and empty (NULL)
l1[[5]] <- "More elements!" # The list l1 had 4 elements, we're adding a 5th here.
l1[[8]] <- 1:11 # We added an 8th element, but not 6th or 7th. Those will be created empty (NULL)
l1$Something <- "A thing"  # Adds a ninth element - "A thing", named "Something"
```

#### Data Frames 
The data frame is a special kind of list used for storing dataset tables. Think of rows as cases, columns as variables. Each column is a vector or factor.

##### Creating a dataframe:
````{r}
dfr1 <- data.frame( ID=1:4,
                    FirstName=c("John","Jim","Jane","Jill"),
                    Female=c(F,F,T,T), 
                    Age=c(22,33,44,55) )

dfr1$FirstName   # Access the second column of dfr1. 
# Notice that R thinks this is a categorical variable 
# and so it's treating it like a factor, not a character vector.
dfr1
````

#### Let's get rid of the factor by telling R to treat FirstName as a vector:
```{r}
dfr1$FirstName <- as.vector(dfr1$FirstName)
```

#### Alternatively, you can tell R you don't like factors from the start using stringsAsFactors=FALSE
````{r}
dfr2 <- data.frame(FirstName=c("John","Jim","Jane","Jill"), stringsAsFactors=FALSE)
dfr2$FirstName   # Success: not a factor.
````

### Access elements of the data frame
````{r}
dfr1[1,]   # First row, all columns
dfr1[,1]   # First column, all rows
dfr1$Age   # Age column, all rows
dfr1[1:2,3:4] # Rows 1 and 2, columns 3 and 4 - the gender and age of John & Jim
dfr1[c(1,3),] # Rows 1 and 3, all columns
```

### Find the names of everyone over the age of 30 in the data
```{r}
dfr1[dfr1$Age>30,2]
```

#### Find the average age of all females in the data:
````{r}
mean ( dfr1[dfr1$Female==TRUE,4] )
```


### Flow Control 
##### if (condition) expr1 else expr2
````{r}
x <- 5; y <- 10
if (x==0) y <- 0 else y <- y/x #  
y
````

#### for (variable in sequence) expr
````{r}
ASum <- 0; AProd <- 1
for (i in 1:x)  
{
  ASum <- ASum + i
  AProd <- AProd * i
}
ASum  # equivalent to sum(1:x)
AProd # equivalemt to prod(1:x)

# while (condintion) expr
while (x > 0) {print(x); x <- x-1;}

# repeat expr, use break to exit the loop
repeat { print(x); x <- x+1; if (x>10) break}
````

### R plots and colors
In most R functions, you can use named colors, hex, or rgb values: (In the simple base plot chart below x and y are point coordiantes, pch is the point symbol shape, cex is the point size, and col is the color. to see the parameters for ploting in base R, check out ?par
````{r}
plot(x=1:10, y=rep(5,10), pch=19, cex=5, col="dark red")
points(x=1:10, y=rep(6, 10), pch=19, cex=5, col="#557799")
points(x=1:10, y=rep(4, 10), pch=19, cex=5, col=rgb(.25, .5, .3))
```

You may notice that rgb here ranges from 0 to 1. While this is the R default, you can also set it for the 0-255 range: 
````{r}
rgb(10, 100, 100, maxColorValue=255) 
```

We can also set the opacity/transparency using the parameter 'alpha' (range 0-1):
````{r}
plot(x=1:5, y=rep(5,5), pch=19, cex=16, col=rgb(.25, .5, .3, alpha=.5), xlim=c(0,6))  
```

If we have a hex color representation, we can set the transparency alpha using 'adjustcolor' from package 'grDevices'. For fun, let's also set the the plot background to gray using the par() function for graphical parameters.
```{r}
par(bg="black")
col.tr <- grDevices::adjustcolor("#557799", alpha=0.7)
plot(x=1:5, y=rep(5,5), pch=19, cex=20, col=col.tr, xlim=c(0,6)) 
par(bg="white")
```

### If you plan on using the built-in color names, here's what they are: 
````{r}
colors()
grep("blue", colors(), value=T)
```

In many cases, we need a number of contrasting colors, or multiple shades of a color. R comes with some predefined palette function that can generate those for us.
````{r}
pal1 <- heat.colors(5, alpha=1)   # generate 5 colors from the heat palette, opaque
pal2 <- rainbow(5, alpha=.5)      # generate 5 colors from the heat palette, semi-transparent
plot(x=1:10, y=1:10, pch=19, cex=10, col=pal1)
plot(x=10:1, y=1:10, pch=19, cex=10, col=pal2)
```

We can also generate our own gradients using colorRampPalette. Note that colorRampPalette returns a *function* that we can use to generate as many colors from that palette as we need.
````{r}
palf <- colorRampPalette(c("gray70", "red")) 
plot(x=10:1, y=1:10, pch=19, cex=10, col=palf(100))
```

To add transparency to colorRampPalette, you need to add a parameter `alpha=TRUE`:
```{r}
palf <- colorRampPalette(c(rgb(1,1,1, .2),rgb(.8,0,0, .7)), alpha=TRUE)
plot(x=10:1, y=1:10, pch=19, cex=10, col=palf(10)) 
```







